% The acknowledgements section is optional. It can cover at most two pages.
\acknowledge{
	% If don't have an acknowledgment, comment out or delete this whole file
    I would not have made it through this journey without the support of many people.
Title: Begin the secVtion with a heading such as "Acknowledgements" or "Acknowledgments" centered at the top of the page.

Introduction: Write a brief introductory paragraph expressing your gratitude and acknowledging the support and contributions of individuals and organizations throughout your research journey.

Advisors and Supervisors: Start by acknowledging your primary dissertation advisor(s) and other members of your dissertation committee who provided guidance and support throughout your research. Mention their names, titles, and the academic institution they belong to.

Funding Agencies and Grants: Acknowledge any funding agencies, organizations, or institutions that provided financial support for your research. Include the names of the grants or scholarships received, if applicable.

Academic and Research Institutions: Recognize any academic institutions, departments, or research centers that provided facilities, resources, or collaborative opportunities during your PhD program.

Colleagues and Peers: Express gratitude to your fellow PhD students, colleagues, and peers who provided assistance, shared knowledge, or engaged in discussions that contributed to your research.

Family and Friends: Acknowledge the support and understanding of your family and friends throughout your doctoral journey. Express your appreciation for their encouragement, patience, and love.

Technical Support and Experts: Thank any technical support staff, laboratory assistants, or experts who assisted you with experiments, data analysis, or any specialized equipment.

Participants and Interviewees: If your research involved human participants or interviews, express your gratitude to those individuals who participated in your study, highlighting their valuable contributions.

Mentors and Teachers: Acknowledge any mentors, teachers, or professors who have played a significant role in shaping your academic and intellectual development.

General Support: Acknowledge any individuals or groups who provided general support, encouragement, or inspiration throughout your doctoral journey.

Conclusion: Conclude the section with a final statement expressing your sincere appreciation for the collective support and guidance you have received during your research endeavors.

With some sense of accomplishment,
I am delighted to present this dissertation, which is the culmination of over five years of dedication to expanding our understanding of biology.
This journey would not have been possible without the support, guidance, and encouragement of a multitude of individuals.
I wish I could thank everyone involved personally, but my memory is not what it used to be when I first embarked on this quest...
As I reflect on this long process, 
I would like to offer my gratitude to those who have been instrumental in shaping my academic growth and in keeping my mental sanity.

I am so fortunate to have found not one, but two incredibly kind, patient and generous advisors.
Andy, thank you for your enthusiasm for science, for supporting my growth as a scientist, and for your encouragement through ups and downs.
Peter, thank your for your generosity and patience, for providing valuable and thorough insights, and for being a pillar of support throughout my academic journey.
I also need to thank both of you for nurturing sense of community and collegiality within our CoLab.

Next, I would like to thank my lab mates, collaborators and colleagues who have provided academic and emotional support throughout my journey.
Thank you to my amazingly helpful, supportive and fun lab mates: Jeff Adrion, CJ Battey, Jared Galloway, Matt Lukac, Anastasia Teterina, Gabby Coffing, Victoria Caudill, Saurabh Belsare, Vince Buffalo, Gilia Patterson, Chris Smith, Clara Rehmann, Jordan Anderson, Nate Pope, Jiseon Min, Jordan Rodriguez, Georgia Tsambos, Silas Tittes, Scott Small and Bruce Edelman.
I would like to thank my committee members, Matt Streisfeld, Kelley Harris, Jayson Paulose, and Patrick Phillips for all their support throughout these years.
My cohort mates have been so wonderful throughout the years and helped me make it through: Sabrina Mostoufi, Sophia Frantz, Monika Ruwaimana, Lina Aoyama, Max Spencer and Kayla Evans. 
I am thankful to many other members of the Institute of Ecology and Evolution, including the behind the scenes staff that helped me with so many travel and purchase needs throughout the years;
thank you Arlene Crain, Sara Nash, Maria Heider, and Leah Frazier.
Thanks are also in order to my always supportive collaborators and members of the tskit ecosystem and the PopSim consortium.

Before my PhD, I was lucky to find amazing people who have helped shape me as a scientist and biologist.
I would like to thank my undergraduate thesis advisor, Fernando Marques, for helping me get my feet wet with academic research.
Rodrigo Cogni, my master's advisor, helped me make the jump into empirical population genetics.
John Pool welcomed me into his lab and gave me incredible freedom to do whatever research I wanted.
Paulo Inácio Prado, Alexandre Adalardo de Oliveira and Adriana Martini opened my eyes to the importance of theory, models and simulation in biology.
The entire crew behind the Atlantic Forest Field Ecology course (offered by the Institute of Biosciences at the Universidade de São Paulo) have helped me become a better naturalist.

I want to wholeheartedly thank my family and close friends,
whose unwavering encouragement and love have sustained me during the demanding years of my academic pursuit.
Thank you to my mom, Kelly, for always supporting me throughout anything I set my mind to.
Her strength and determination were passed down to me, and I would not have made it through without it.
Thank you to my dad, Aguinaldo, for showing me the great things one can accomplish with hard work and grit.
I am proud for what my parents accomplished in raising me and my siblings
at such a young age with little family support and no college education.
I also need to acknowledge my parents for partially supporting me financially throughout my entire academic career,
as this journey would have been orders of magnitude harder without it.
I would to thank my siblings, Giovanna and João, for being a source of love, support and laughter.
Lastly, I need to thank my partner Luiza.



}

