\school{University of Oregon, Eugene, OR, USA}
\school{Universidade de São Paulo, São Paulo, SP, Brasil}

\degree{Doctor of Philosophy, Biology, 2024, University of Oregon}
\degree{Master of Science, Genetics and Evolutionary Biology, 2018, Universidade de São Paulo}
\degree{Bachelor of Science, Biology, 2015, Universidade de São Paulo}

\interests{Evolutionary Biology}
\interests{Population Genetics}
\interests{Computational Biology}

\award{Harvey E. Lee Graduate Scholarship, University of Oregon, 2022-2023}
\award{Marthe E. Smith Memorial Science Scholarship, University of Oregon, 2022-2023}
\award{Hill Fund Graduate Award, University of Oregon, 2020-2021}
\award{Genetics Training Grant, University of Oregon, 2019-2021}
\award{Research Internship Abroad Fellowship, The São Paulo Research Foundation, 2017-2018}
\award{Master's Research Fellowship, The São Paulo Research Foundation, 2016-2018}
\award{Undergraduate Research Fellowship, The São Paulo Research Foundation, 2013-2014}
\award{} % extra empty award is necessary

% Example publications:
\publication{Rodrigues, M. F., Kern, A. D., \& Ralph, P. L. (2024).
             Shared evolutionary processes shape landscapes of genomic 
             variation in the great apes. \emph{Genetics}, iyae006.}
\publication{Estevez-Castro, C.F., Rodrigues, M.F., Babarit, A.,
             ... \& Olmo, R. P. (2024). Neofunctionalization driven by
             positive selection led to the retention of the loqs2 gene
             encoding an \emph{Aedes} specific dsRNA binding protein. 
             \emph{BMC Biology} 22, s12915-024-01821-4.}
\publication{Lauterbur, M. E., Cavassim, M. I. A., Gladstein, A. L.,
             Gower, G., Pope, N. S., Tsambos, G., ..., Rodrigues, M. F., ... \& Gronau, I. (2023).
             Expanding the stdpopsim species catalog, and lessons learned 
             for realistic genome simulations. \emph{eLife}, 12, RP84874.}
\publication{Baumdicker, F., Bisschop, G., Goldstein, D., Gower, G., 
             Ragsdale, A. P., Tsambos, G., ..., Rodrigues, M. F., ... \& Kelleher, J. (2022).
             Efficient ancestry and mutation simulation with msprime 1.0.
             \emph{Genetics}, 220(3), iyab229.}
\publication{Rodrigues, M. F., Vibranovski, M. D., \& Cogni, R. (2021). 
             Clinal and seasonal changes are correlated in 
             \textit{Drosophila melanogaster} natural populations. 
             \emph{Evolution}, 75(8), 2042-2054.}
\publication{Rodrigues, M. F., \& Cogni, R. (2021). Genomic responses to 
             climate change: Making the most of the \emph{Drosophila} model. 
             \emph{Frontiers in Genetics}, 12, 676218.}
\publication{Stankowski, S., Chase, M. A., Fuiten, A. M., Rodrigues, M. F., 
             Ralph, P. L., \& Streisfeld, M. A. (2019). Widespread selection
             and gene flow shape the genomic landscape during a radiation of 
             monkeyflowers. \emph{PLoS Biology}, 17(7), e3000391.}
\publication{} % extra empty publication is necessary
